Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Clean up and revision of Introduction.

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments as an integral part of
successful research teams. The WSSSPE workshop  provided a forum
for discussion of the challenges around scientific software, including both positions and experiences. The
workshop short papers and debates have been archived to provide a basis for continued
discussion, and have fed into the collaborative writing of this
document. Additional papers based on extended versions of workshop
submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved
an estimated~90 to~150 participants during the day has led to further activities:
Two additional WSSSPE workshops are planned  (at the~2014 SciPy and SC14 conferences),
funds to support the workshops have been obtained from
the National Science Foundation and the Gordon and Betty Moore
Foundation, and the
original workshop website has been turned into a community website for
further discussion and progress. Additionally, a minisymposium  to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the~2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science'' (SIAM AN14, \url{http://meetings.siam.org}).

This paper summarizes the  workshop papers
and discussion, and identifies issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations~(\S\ref{sec:keynotes}), the three
panels~(\S\ref{sec:devel}-\ref{sec:community}),
cross-cutting issues~(\S\ref{sec:cross-cutting}),
use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).
Additionally, the accepted workshop papers are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
in Appendix~\ref{sec:attendees}.

>>>>

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments as an integral part of
successful research teams. This first WSSSPE workshop  provided a forum
for discussion of the challenges around sustaining scientific software, including
contributed short papers in the form of both positions and experience
reports. These short papers, as well as notes from the debates around
them, have been archived to provide a basis for continued
discussion, and have fed into the collaborative writing of this
report.  Some of the workshop submissions have been extended to full
papers, which form part of the same special journal edition in which this report
appears. The workshop generated a high level of interest; an
estimated~90 to~150 participants were in attendance at different times
of the day. The interest and discussions have already led to follow-up activities:
A smaller Python-specific workshop is planned to be held at the~2014
SciPy conference, and a follow-on WSSSPE2 workshop has been proposed
to, and meanwhile been
accepted for the SC14 conference. In addition, 
funds to support the workshops have been obtained from
the US National Science Foundation (NSF) and the Gordon and Betty Moore
Foundation, and the
original workshop website has been turned into a community website to engender
further discussion and progress. Additionally, a minisymposium  at the~2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science'' (SIAM AN14, \url{http://meetings.siam.org})  is being co-organized by a
WSSSPE1 participant to
further explore some of the key issues raised at the workshop.

This report attempts to summarize the various aspects of the
workshop.  The remainder of this paper first gives an overview of the
process with which the workshop was put together
(\S\ref{sec:process}),  then proceeds with summaries of the two
keynote presentations~(\S\ref{sec:keynotes}), followed by summaries of
the workshop papers grouped by the three thematic workshop panels to which they
were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two broader
cross-cutting issues surfaced repeatedly, and are
discussed separately~(\S\ref{sec:cross-cutting}), as are
use cases for sustainable software~(\S\ref{sec:use-cases}). The
summaries are based not only on the papers and panel presentations, but
also on the many comments raised in both the onsite and online discussions
that accompanied the workshop elements, as documented by collaborative
note taking during the workshop~\cite{WSSSPE1-google-notes}. We
conclude with issues and lessons learned, as well as plans for future activities~(\S\ref{sec:conclusions}).
The short papers accepted to the workshop are listed in
Appendix~\ref{sec:papers}, and a partial list of workshop attendees
can be found in Appendix~\ref{sec:attendees}.
